Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-28 paper.tex
minor update
major restructuring
notation change, begun adiabatic notes
more adiabatic discussion, plus a new fig of growth rates
write up edits
minor updates
extended analytic discussion to thermal relaxation cases
new fig for analytic discussion with therm relax
adding applications
writing summary now
completed rough summary
drafted summary
abstract added, intro left to do

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\zmax}{Z_s}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

\title{Linear vertical shear instability in astrophysical disks: a parameter study}

\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu}

\begin{abstract}

\end{abstract}

\section{Introduction}

\input{setup}
\input{linear_iso}
%\input{results}
%\input{results_gi}
%\input{results_azi} 
%\input{summary}

%\acknowledgements
%\appendix
%\input{appendix}

>>>>

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

\title{Linear analysis of the vertical shear instability in astrophysical disks}

\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is necessary to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks, including an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency. We explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicate that the VSI
  requires a short thermal relaxation timscale. For vertically
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature (which gives rise to vertical
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular mometum transport is the
central problem in the theory of accretion disks
\citep{lydenbell74}. The
magneto-rotational instability in the context of astrophysical disks 
\citep{balbus91} is now accepted as a robust pathway to
magneto-hydrodynamic (MHD) turbulence, which transports angular momentum
outwards and thus enables mass accretion.

%stuff about vsi

%recent calculations (nelson, mcnally)

%why linear approach? computationally fast (need very high resolution) 
%can be used as a first estimate

%what we're doing

%paper organization

\input{setup}
\input{linear_iso}
\input{linear_adia}
\input{numerical}
\input{application} 
\input{summary}

%\acknowledgements
\appendix
\input{appendix}

### 2015-01-28 summary.tex
abstract added, intro left to do

We briefly studied vertically non-isothermal disks using our linear
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to thermal relaxation. When a
stable stratification is introduced, the dependence of VSI growth
rates on the thermal relaxation timescale is qualitatively similar to
that observed for vertically isothermal disks.

>>>>

We briefly studied vertically non-isothermal disks using our linear
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to thermal relaxation. When a
stable stratification is introduced, the reduction of VSI growth
rates by a finite thermal relaxation timescale is qualitatively similar to
that observed for vertically isothermal disks.
